\mpx{}: mixed precision Training for JAX

### 2025-05-28 doc/paper/main.tex
temp

\usepackage{caption}
\usepackage{subcaption}

@@

\newcommand{\mpx}{\textsc{MPX}}

@@

\begin{document}

\maketitle

@@

\begin{figure}[t]
    \centering
    \caption{Transformation of training pipelines via MPX. An existing training pipeline (Top figure) can be transformed, by applying \texttt{@mpx.filter\_grad} to the forward pass. It automatically casts all inputs to half precision and calculates the gradient via dynamic loss scaling.}
    \label{fig:overview}
    \begin{subfigure}[t]{0.99\textwidth}
        \centering
        \caption{Full Precision Training via Equinox}
        \includegraphics[width=0.99\textwidth]{assets/EQXOverview.pdf}
    \end{subfigure}\hspace{0.1cm}
    \begin{subfigure}[t]{0.99\textwidth}
        \centering
        \caption{Half Precision Training via Equinox and \mpx{}}
        \includegraphics[width=0.99\textwidth]{assets/MPXOverview.pdf}
    \end{subfigure}\hspace{0.1cm}
\end{figure}

@@

\begin{example}[h]
    \centering
    \caption{Implementation of multi-headed self-attention via MPX. For the largest part, the model can be implemented in common toolboxes (here, Equinox~\cite{kidger2021equinox}). Only critical function that should be implemented in full precision, like softmax or layernorm (due to mean and std calculations), have to be transformed via \mpx{}.}
    \label{ex:update}
    \begin{subfigure}[t]{0.98\textwidth}
        \centering
        \caption{Full Precision via Equinox}
        \vspace{-0.4cm}
        \begin{mdframed}[
            backgroundcolor=white!10!black,
            leftmargin=0.5cm,hidealllines=true,%
            innerleftmargin=0.2cm,innerrightmargin=0.2cm,innertopmargin=0cm,innerbottommargin=-0.10cm
        ]
            % \fontsize{8pt}{8pt}\selectfont
            \begin{lstlisting}[, escapeinside={<@}{@>},language=Python, basicstyle=\ttfamily\linespread{1.1}\footnotesize\color{white!70!black}, frame=l, label={lst:model}, keywordstyle=\color{blue!30!white}, commentstyle=\color{gray}, stringstyle=\color{orange!80!black}, numbers=left, backgroundcolor=\color{white!10!black}, numbers=left, framesep=0.8cm, framerule=0pt,  xleftmargin=0.7cm,]
grads = eqx.filter_grad(loss)(model, batch)
updates, optimizer_state = optimizer.update(
    grads, optimizer_state, eqx.filter(model, eqx.is_array))
model = eqx.apply_updates(model, updates)
            \end{lstlisting}
        \end{mdframed}
    \end{subfigure}\hspace{0.1cm}
    \begin{subfigure}[t]{0.98\textwidth}
        \centering
        \vspace{0.2cm}
        \caption{mixed precision via MPX}
        \vspace{-0.4cm}
        \begin{mdframed}[
            backgroundcolor=white!10!black,
            leftmargin=0.5cm,hidealllines=true,%
            innerleftmargin=0.2cm,innerrightmargin=0.2cm,innertopmargin=0cm,innerbottommargin=-0.10cm
        ]
            % \fontsize{8pt}{8pt}\selectfont
            \begin{lstlisting}[, escapeinside={<@}{@>},language=Python, basicstyle=\ttfamily\linespread{1.1}\footnotesize\color{white!70!black}, frame=l, label={lst:model}, keywordstyle=\color{blue!30!white}, commentstyle=\color{gray}, stringstyle=\color{orange!80!black}, numbers=left, backgroundcolor=\color{white!10!black}, numbers=left, framesep=0.8cm, framerule=0pt,  xleftmargin=0.7cm,]
loss_scaling, grads_finite, grads = mpx.filter_grad(loss, loss_scaling)(
    model, batch)
model, optimizer_state = mpx.optimizer_update(
    model, optimizer, optimizer_state, grads,grads_finite)
            \end{lstlisting}
        \end{mdframed}
    \end{subfigure}

\end{example}

@@

\begin{example}[h]
    \centering
    \begin{minipage}{0.95\linewidth}
            \caption{Implementation of model updates via Equinox and \mpx{}. \mpx{} is designed such that its model update routine closely resembles the one of Equinox.}
            \label{ex:model}
        \begin{mdframed}[
    backgroundcolor=white!10!black,
    leftmargin=0.5cm,hidealllines=true,%
  innerleftmargin=0.2cm,innerrightmargin=0.2cm,innertopmargin=0cm,innerbottommargin=-0.10cm
]
        % \fontsize{8pt}{8pt}\selectfont
            \begin{lstlisting}[, escapeinside={<@}{@>},language=Python, basicstyle=\ttfamily\linespread{1.1}\footnotesize\color{white!70!black}, frame=l, label={lst:model}, keywordstyle=\color{blue!30!white}, commentstyle=\color{gray}, stringstyle=\color{orange!80!black}, numbers=left, backgroundcolor=\color{white!10!black}, numbers=left, framesep=0.8cm, framerule=0pt,  xleftmargin=0.7cm,]
class MultiHeadAttentionBlock(eqx.Module):
    dense_qs: eqx.nn.Linear
    dense_ks: eqx.nn.Linear
    dense_vs: eqx.nn.Linear
    dense_o: eqx.nn.Linear
    num_heads: int
    layer_norm: eqx.nn.LayerNorm

@@

        return outputs
        \end{lstlisting}
        \end{mdframed}
    \end{minipage}
\end{example}

>>>>

\usepackage{microtype}

\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\usepackage{caption}
\usepackage{subcaption}
\captionsetup[subfigure]{justification=centering}

@@

\newcommand{\mpx}{\textsc{MPX}}
\newcommand{\capt}[1]{\mdseries{\emph{#1}}}

@@

\begin{document}
\sloppy
\maketitle

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[t]{0.99\textwidth}
        \centering
        \includegraphics[width=0.99\textwidth]{assets/EQXOverview.pdf}
        \caption{Full Precision Training via Equinox}
        \vspace{0.2cm}
    \end{subfigure}\hspace{0.1cm}
    \begin{subfigure}[t]{0.99\textwidth}
        \centering
        \includegraphics[width=0.99\textwidth]{assets/MPXOverview.pdf}
        \caption{Half Precision Training via Equinox and \mpx{}}
    \end{subfigure}\hspace{0.1cm}
    \caption{Transformation of training pipelines via MPX. \capt{An existing training pipeline (Top figure) can be transformed, by applying \texttt{@mpx.filter\_grad} to the forward pass. It automatically casts all inputs to half precision and calculates the gradient via dynamic loss scaling and \texttt{eqx.filter\_grad}.}}
    \label{fig:overview}
\end{figure}

@@

\begin{example}[h]
    \centering
    \begin{subfigure}[t]{0.98\textwidth}
        \centering
        \begin{mdframed}[
            backgroundcolor=white!10!black,
            leftmargin=0.5cm,hidealllines=true,%
            innerleftmargin=0.2cm,innerrightmargin=0.2cm,innertopmargin=0cm,innerbottommargin=-0.10cm
        ]
            % \fontsize{8pt}{8pt}\selectfont
            \begin{lstlisting}[, escapeinside={<@}{@>},language=Python, basicstyle=\ttfamily\linespread{1.1}\footnotesize\color{white!70!black}, frame=l, label={lst:model}, keywordstyle=\color{blue!30!white}, commentstyle=\color{gray}, stringstyle=\color{orange!80!black}, numbers=left, backgroundcolor=\color{white!10!black}, numbers=left, framesep=0.8cm, framerule=0pt,  xleftmargin=0.7cm,]
grads = eqx.filter_grad(loss)(model, batch)
updates, optimizer_state = optimizer.update(
    grads, optimizer_state, eqx.filter(model, eqx.is_array))
model = eqx.apply_updates(model, updates)
            \end{lstlisting}
        \end{mdframed}
        \vspace{-0.5cm}
        \caption{Full Precision via Equinox}
        \vspace{0.5cm}
    \end{subfigure}\hspace{0.1cm}
    \begin{subfigure}[t]{0.98\textwidth}
        \centering
        \begin{mdframed}[
            backgroundcolor=white!10!black,
            leftmargin=0.5cm,hidealllines=true,%
            innerleftmargin=0.2cm,innerrightmargin=0.2cm,innertopmargin=0cm,innerbottommargin=-0.10cm
        ]
            % \fontsize{8pt}{8pt}\selectfont
            \begin{lstlisting}[, escapeinside={<@}{@>},language=Python, basicstyle=\ttfamily\linespread{1.1}\footnotesize\color{white!70!black}, frame=l, label={lst:model}, keywordstyle=\color{blue!30!white}, commentstyle=\color{gray}, stringstyle=\color{orange!80!black}, numbers=left, backgroundcolor=\color{white!10!black}, numbers=left, framesep=0.8cm, framerule=0pt,  xleftmargin=0.7cm,]
loss_scaling, grads_finite, grads = mpx.filter_grad(loss, loss_scaling)(
    model, batch)
model, optimizer_state = mpx.optimizer_update(
    model, optimizer, optimizer_state, grads,grads_finite)
            \end{lstlisting}
        \end{mdframed}
        \vspace{-0.5cm}
        \caption{mixed precision via MPX}
    \end{subfigure}
    \caption{Implementation of model updates via Equinox and \mpx{}. \mpx{} is designed such that its model update routine closely resembles the one of Equinox.}
    \label{ex:update}
\end{example}

@@

\begin{example}[h]
    \centering
    \begin{minipage}{0.95\linewidth}
    \begin{mdframed}[
    backgroundcolor=white!10!black,
    leftmargin=0.5cm,hidealllines=true,%
  innerleftmargin=0.2cm,innerrightmargin=0.2cm,innertopmargin=0cm,innerbottommargin=-0.10cm
]
        % \fontsize{8pt}{8pt}\selectfont
            \begin{lstlisting}[, escapeinside={<@}{@>},language=Python, basicstyle=\ttfamily\linespread{1.1}\footnotesize\color{white!70!black}, frame=l, label={lst:model}, keywordstyle=\color{blue!30!white}, commentstyle=\color{gray}, stringstyle=\color{orange!80!black}, numbers=left, backgroundcolor=\color{white!10!black}, numbers=left, framesep=0.8cm, framerule=0pt,  xleftmargin=0.7cm,]
class MultiHeadAttentionBlock(eqx.Module):
    dense_qs: eqx.nn.Linear
    dense_ks: eqx.nn.Linear
    dense_vs: eqx.nn.Linear
    dense_o: eqx.nn.Linear
    num_heads: int
    layer_norm: eqx.nn.LayerNorm

@@

        return outputs
        \end{lstlisting}
        \end{mdframed}
    \end{minipage}
    \caption{Implementation of multi-headed self-attention via MPX. \capt{For the largest part, the model can be implemented in common toolboxes (here, Equinox~\cite{kidger2021equinox}). Only critical function that should be implemented in full precision, like softmax or layernorm (due to mean and std calculations), have to be transformed via \mpx{}.}}
    \label{ex:model}
\end{example}

\section{Evaluation}
We evaluate \mpx{} on two different machines: A mid-level desktop PC (AMD Ryzen 9 5950X, \qty{82}{\giga\byte} RAM, Nvidia GeForce RTX4070 GPU with \qty{12}{\giga\byte} VRAM) and a node on a high-performance cluster (CLAIX-2023, Intel Xeon 8468 Sapphire CPUs, \qty{512}{\giga\byte} RAM, four NVIDIA H100 GPUs with \qty{96}{\giga\byte} VRAM each connected via NVLink).
On paper, the desktop PC's GPU half precision computation speed is the same for half and full precision, while the H100 GPUs has double the speed for half than for full precision.

We train small vision transformers (ViT)~\cite{dosovitskiy2020image,steiner2021train}.
On both machines, the transformer has a feature size is 256, the residual blocks have one hidden layer with 800 neurons.
The desktop PC trains the ViT on Cifar100 and the high-performance cluster on ImageNet1k, equally dividing each batch between the GPUs using a data-parallel approach.
We train without gradient accumulation, which would be another way to optimize runtime and memory consumption.
We vary the number of batches and measure the time the machines need to for each training step on the entire dataset, excluding the times used for data loading.

We measure the RAM the program uses by turning off automatic preallocation and setting ``XLA\_PYTHON\_CLIENT\_PREALLOCATE'' to ``platform'', which minimizes the VRAM consumption of JAX\footnote{\url{https://docs.jax.dev/en/latest/gpu_memory_allocation.html}}
For training speed, we turn both off.

\begin{figure}[h]
    \centering
    \begin{subfigure}[t]{0.49\linewidth}
    \begin{tikzpicture}
        \begin{axis}[
            width=0.8\textwidth,
            height=7cm,
            xlabel={Batch Size},
            ylabel={VRAM (\qty{}{\giga\byte})},
            legend pos=south east,
            grid=major,
            grid style={dashed,gray!30},
            xmin=0,
            ymin=0
        ]
        % First graph: Full Precision
        \addplot[
            color=blue,
            mark=o,
            thick
        ] coordinates {
            (128, 2.290)
            (256, 4.724)
            (512, 8.860)
        };
        \addlegendentry{Full Precision}

        % Second graph: Mixed Precision
        \addplot[
            color=red,
            mark=square*,
            thick,
            dashed
        ] coordinates {
            (128, 1.52)
            (256, 2.418)
            (512, 4.900)
            (768, 7.054)
            (1024, 9.222)
        };
        \addlegendentry{Mixed Precision}
        \end{axis}
    \end{tikzpicture}
    \caption{Desktop Computer (RTX4070)}
    \end{subfigure}
    \begin{subfigure}[t]{0.49\linewidth}
    \begin{tikzpicture}
        \begin{axis}[
            width=0.8\textwidth,
            height=7cm,
            xlabel={Batch Size},
            ylabel={VRAM (\qty{}{\giga\byte})},
            legend pos=south east,
            grid=major,
            grid style={dashed,gray!30},
            xmin=0,
            ymin=0
        ]
        % First graph: Full Precision
        \addplot[
            color=blue,
            mark=o,
            thick
        ] coordinates {
            (128, 2.290)
            (256, 4.724)
            (512, 8.860)
        };
        \addlegendentry{Full Precision}

        % Second graph: Mixed Precision
        \addplot[
            color=red,
            mark=square*,
            thick,
            dashed
        ] coordinates {
            (128, 1.52)
            (256, 2.418)
            (512, 4.900)
            (768, 7.054)
            (1024, 9.222)
        };
        \addlegendentry{Mixed Precision}
        \end{axis}
    \end{tikzpicture}
    \caption{High-Performance Cluster ($4\times$ H100)}
    \end{subfigure}
    \caption{Comparison of GPU VRAM consumed for full precision and mixed precision as a function of the number of batches.}
    \label{fig:memory}
\end{figure}

\begin{figure}[h]
    \centering
    \begin{subfigure}[t]{0.49\linewidth}
    \begin{tikzpicture}
        \begin{axis}[
            width=0.8\textwidth,
            height=7cm,
            xlabel={Batch Size},
            ylabel={Training Time per Epoch (s)},
            legend pos=south east,
            grid=major,
            grid style={dashed,gray!30},
            xmin=0,
            ymin=0,
        ]
        % First graph: Full Precision
        \addplot[
            color=blue,
            mark=o,
            thick
        ] coordinates {
            (128, 19.7)
            (256, 18.8)
            (512, 19.7)
        };
        \addlegendentry{Full Precision}

        % Second graph: Mixed Precision
        \addplot[
            color=red,
            mark=square*,
            thick,
            dashed
        ] coordinates {
            (128, 12.6)
            (256, 12.0)
            (512, 11.4)
            (768, 11.6)
            (1024, 11.5)
        };
        \addlegendentry{Mixed Precision}
        \end{axis}
    \end{tikzpicture}
    \caption{Desktop Computer (RTX4070)}
    \end{subfigure}
    \begin{subfigure}[t]{0.49\linewidth}
    \begin{tikzpicture}
        \begin{axis}[
            width=0.8\textwidth,
            height=7cm,
            xlabel={Batch Size},
            ylabel={Training Time per Epoch (s)},
            legend pos=south east,
            grid=major,
            grid style={dashed,gray!30},
            xmin=0,
            ymin=0,
        ]
        % First graph: Full Precision
        \addplot[
            color=blue,
            mark=o,
            thick
        ] coordinates {
            (128, 19.7)
            (256, 18.8)
            (512, 19.7)
        };
        \addlegendentry{Full Precision}

        % Second graph: Mixed Precision
        \addplot[
            color=red,
            mark=square*,
            thick,
            dashed
        ] coordinates {
            (128, 12.6)
            (256, 12.0)
            (512, 11.4)
            (768, 11.6)
            (1024, 11.5)
        };
        \addlegendentry{Mixed Precision}
        \end{axis}
    \end{tikzpicture}
    \caption{High-Performance Cluster ($4\times$ H100)}
    \end{subfigure}
    \caption{Comparison of training step times for full precision and mixed precision as a function of the number of batches.}
    \label{fig:timing}
\end{figure}

Mixed precision training via \mpx{} reduces both RAM and processing times.
For the desktop PC, the RAM is reduced by factor $1.8\times$ and processing time by $1.7\times$.
The speedup is caused by reduced memory loading times as the RTX4070 features no computing speedup for half precision.

\section{Conclusion}

\mpx{} offers an easy-to-use solution for mixed-precision training in JAX.
It is able to handling arbitrary PyTrees and provides a seamless integration with popular libraries like Equinox and Flax. 
Through dynamic loss scaling and precision casting, MPX ensures efficient memory usage and accelerates training without compromising model accuracy. 
The evaluation demonstrates significant reductions in RAM usage and processing times, particularly on hardware that supports half-precision computation. 
Overall, MPX presents a compelling advancement in optimizing large-scale neural network training pipelines while maintaining flexibility and ease of use.
